Binarity's Biases on Transit Survey Occurrence Rates

### 2018-02-24 paper/main.tex
text improvements

Here, though, we will focus exclusively on problems that arise from
the fact that many stars exist in gravitationally bound binary
systems.  We will also focus on the errors in planet occurrence as a
function of radius, rather than orbital period. This is because when
more than one transit is detected (as is usually required by the
surveyors), the orbital periods can be measured without ambiguity
regardless of whether the host star is single or one member of a
binary.  Even with this narrow focus, there are numerous sources of
error.  All three of the quantities in
Equation~\ref{eq:occ_rate_simple} are biased:
\begin{enumerate}
%    
    \item The number of detected planets, $N_{\det}$, is actually the
      number of detected planets that {\it appear} to have size $r$,
      orbiting stars that {\it appear} to have mass $M$.  Whenever the
      planet-hosting star is part of a binary,
%    
    \begin{itemize}
        \item the planet's size could be misclassified because of the
          reduction in the amplitude of the photometric signal;
%        
        \item the host star's properties could be misclassified
          because its light is combined with a second star of a
          different spectral type.
%        
    \end{itemize}
%    
    \item The number of searchable stars, $N_\star$, is biased
%    
    \begin{itemize}
%        
        \item toward lower values, because it does not include all of
          the secondary stars that were inadvertently searched for
          transiting planets;
%        
        \item toward higher values, because some of the stars that
          appeared to be searchable are in fact binaries for which the
          amplitude of the photometric signal would have been reduced
          to an undetectable level.
%        
    \end{itemize}
%    
    \item The transit probability $p_{\rm tra}$ is biased because the
      planet-hosting star could be misclassified.  At fixed orbital
      period, the transit probability is proportional to
      $\rho^{-1/3}$, where $\rho$ is the stellar mean
      density~\citep{Winn2010}.  Therefore, any errors in determining
      the host star's mean density lead to errors in the correction
      for the transit probability.
%    
\end{enumerate}

@@

Since the effects of binarity are most pronounced when the two stellar
components are similar, we begin by considering a universe in which
all stars are identical, with mass $M$, radius $R$, and luminosity
$L$.  Single stars are uniformly distributed in space with a number
density of $n\s$ stars per cubic parsec, and binaries are uniformly
distributed with number density $n\b$.  In this scenario, stars are
never misclassified because the combined light of a binary has the
same color and spectrum as a single star.  We further assume that all
planets have the same radius, $r$, and occur around single stars and
members of binaries at the same rate, $\Lambda(r)$.

Our naive observers conduct a transit survey. To calculate the
occurrence rate of planets with radius $r$, they count the number of
detections of signals with amplitude $(r/R)^2$.  Then they identify
all the sources that appear to have been searchable for a signal of
amplitude $(r/R)^2$.  Here and throughout the rest of this paper, we
assume that the limiting source of noise is the photon-counting noise
from the source, i.e., $\sigma \propto 1/\sqrt{F}$, where
where $F$ is the total flux of the source.
Thus the observers determine the minimum $F_0$ for which detection would have been
possible, and count the number of sources with $F_{\rm tot}>F_0$. This will
include all the single stars out to a maximum distance
\begin{equation}
  d_0 = \sqrt{\frac{L}{4\pi F_0}}.
  \label{eq:dmax}
\end{equation}
Since binaries are twice as luminous, the condition $F>F_0$ will
include binaries out to the larger distance of $d_0\sqrt{2}$.
None of the stars in binaries will
appear to have a planet of radius $r$, because of the dilution of
the transit signal.  Thus, the {\it apparent} occurrence rate $\Lambda\a$ of
planets of radius $r$ is
\begin{equation}
    \Lambda\a(r) = 
        \frac{\Lambda(r) \, n\s d_0^3}
        {n\s d_0^3 + n\b (d_0\sqrt{2})^3}.
    \label{eq:rate_at_rp}
\end{equation}
This apparent rate is larger than the true rate by a factor
\begin{equation}
    \frac{\Lambda\a(r)}{\Lambda(r)} = 
        \frac{1}{1 + 2^{3/2}(n\b/n\s)}.
    \label{eq:correction_rp}
\end{equation}

The observers will also detect some transiting planets around stars
with binary companions.  The amplitude of these signals is
\begin{equation}
  \frac{L(r/R)^2}{L + L} = \frac{1}{2}\left(\frac{r}{R}\right)^2 =
  \left(\frac{r/\sqrt{2}}{R}\right)^2,
    \label{eq:delta_obs_general} 
\end{equation}
leading the astronomers to believe they have discovered a population
of planets with radius $r/\sqrt{2}$.  To calculate the corresponding
occurrence rate, they count the stars for which this type of signal
would have been detectable.  The limiting flux for detection in this
case is $4F_0$, because the signal amplitude hase been reduced by a
factor of two and the noise level must also be reduced by a factor of
two.  The condition $F>4F_0$ is met for single stars within a distance
$d_0/2$, and binaries within a distance $d_0\sqrt{2}/2$.  Therefore,
the observers will calculate the occurrence rate of this new type of
planet to be
\begin{equation}
    \Lambda\a\left( \frac{r}{\sqrt{2}} \right) = 
        \frac{2\Lambda(r) \, n\b (d_0\sqrt{2}/2)^3}
        {n\s (d_0/2)^3 + n\b (d_0\sqrt{2}/2)^3}
    =
    \frac{2\Lambda(r)\cdot2^{3/2}(n\b/n\s)}{1 + 2^{3/2} (n\b/n\s)}.
    \label{eq:correction_diluted_rp}
\end{equation}
Figure~\ref{fig:model_1_volumes} illustrates the
volumes enclosing the apparently searchable single and binary stars.

@@

From within this sample of apparently searchable sources, the
observers sum the number of detections $dN_{\rm det}$ of planets
with an apparent radius between $r\a$ and $r\a + dr\a$.  There will be
three contributions to this sum,
\begin{align}
  {\rm d}N_{{\rm det},0} &= 
      \frac{N_\star(r\a) p_{\rm tra}}{1+\mu}
      \Gamma_0(r\a)\mathrm{d}r\a,
  \label{eq:n0} \\
  {\rm d}N_{{\rm det},1} &= 
      \frac{\mu N_\star(r\a) p_{\rm tra}}{1+\mu} 
      \Gamma_1(\sqrt{2}r\a)\mathrm{d}(\sqrt{2}r\a),
	\label{eq:n1} \\
  {\rm d}N_{{\rm det},2} &=
      \frac{\mu N_\star(r\a) p_{\rm tra}}{1+\mu} 
      \Gamma_2(\sqrt{2}r\a)\mathrm{d}(\sqrt{2}r\a).
	\label{eq:n2}
\end{align}
where here and elsewhere, the subscript 0 refers to single stars, 1
refers to primary stars of binaries, and 2 refers to secondary stars
of binaries. The apparent occurrence rate density is therefore
\begin{equation}
  \Gamma\a(r\a) = \frac{{\rm d}N_{\rm det}}{{\rm d}r\a} \frac{1}{N_\star(r\a) p_{\rm tra}} =
  \frac{\Gamma_0(r\a)}{1+\mu} +
  \frac{\mu\sqrt{2}\,\Gamma_1(\sqrt{2}r\a)}{1+\mu} +
  \frac{\mu\sqrt{2}\,\Gamma_2(\sqrt{2}r\a)}{1+\mu}.
	\label{eq:formula_twin_binary}
\end{equation}
Our earlier results, Equations~\ref{eq:rate_at_rp}
and~\ref{eq:correction_diluted_rp}, are recovered by integrating this formula
over $r$ after inserting
\begin{equation}
  \Gamma_i(r) = \Lambda(r\p)\, \hat{\delta}(r-r\p),~~{\rm for}\ i\in\{0,1,2\}
\end{equation}
where $\hat{\delta}$ is the Dirac delta function, and $r\p$ is the
single planet size considered in Section~\ref{sec:model_1}.

@@

The most obvious aspect of the results is that the occurrence rate
density of planets smaller than the breakpoint radius of 2\,$R_\oplus$
is now substantially overestimated.  If secondary stars have the same
planet population as the single stars, then the occurrence rate
density of small planets is overestimated by $50\%$.  The magnitude of
the systematic error decreases if there are fewer planets around
secondaries; if secondaries host no planets at all, the apparent rate
is overestimated by only $10\%$.

The effects on larger planets are not as substantial.  To be
quantitative we consider the occurrence rate of giant planets with
$r>8\,R_\oplus$, by integrating the occurrence rate density displayed
in Figure~\ref{fig:occ_rate_model_3_log}:
\begin{equation}
    \Lambda_{\rm giant,a} =
      \int_{8\,R_\oplus}^{\infty} \Gamma\a(r\a)\,{\rm d}r\a,~~{\rm
      and}~~
    \Lambda_{\rm giant,0} =
      \int_{8\,R_\oplus}^{\infty} \Gamma_0(r)\,{\rm d}r.
\end{equation}
In this case the occurrence rate density is underestimated.  The
difference between the apparent and true rates is largest when the
secondary stars do not host any planets, and has a magnitude
$\Lambda_{{\rm g},0}/\Lambda_{\rm g,a} = 1.13$.
%occ_rate_vs_radius_model_3_withtext_Zsub2_0.00_rpu_22.5.pdf
If instead the secondary stars have half as many giant planets as single
stars, then the factor is reduced from 1.13 to 1.06.

@@

The occurrence calculations that led to this discovery, like almost
all occurrence calculations, did not carefully account for
unresolved binaries.  Intuitively we expect the effect of unresolved
binaries to have a blurring effect on the radius distribution, filling
in any gaps in the radius distribution and making them appear less
empty than in reality.  To investigate the quantitative effects we
make identical assumptions as in Section~\ref{sec:model_3}, except
that the radius distribution is assumed to have a complete absence of
planets with sizes between 1.5 and 2\,$R_\oplus$:
\begin{align}
    f(r)
    &\propto
    \left.
    \begin{cases}
        r^\gamma & \text{for } r\geq 2\,R_\oplus, \\
        0 & \text{for } 1.5\,R_\oplus < r < 2\,R_\oplus, \\
        {\rm constant} & \text{for } r\leq 1.5\,R_\oplus.
    \end{cases}
    \right.
    \label{eq:model4_radius_distribution}
\end{align}
Figure~\ref{fig:model_4} shows the resulting true and apparent occurrence
rate densities.
For the case $Z_2/Z_0=1$, unresolved binaries reduces the contrast
of the gap by nearly a factor of two, while also producing
spurious features for apparent sizes beneath 1.5\,$R_\oplus$.
Thus the gap identified by \citet{fulton_california-_2017} may be
even more devoid of planets than it appears.

@@

The planet population may include some special members with a distinct
radius distribution.  For example in the recent study
by~\citet{petigura_CKS_2017}, hot Jupiters appear as an island in
period-radius space, rather than as a component of a power-law
distribution extending to smaller planets and longer periods.
For this reason we test the effects of unresolved binaries
on a Gaussian radius distribution,
\begin{equation}
    f(r) \propto \exp \left[ -\frac{(r-\bar{r})^2}{2\sigma_r^2} \right],
\end{equation}
with $\bar{r} = 14\,R_\oplus$ and $\sigma_r = 2\,R_\oplus$.  As before,
we allow for different normalizations of the planet populations around
single, primary, and secondary stars: $\Gamma_i(r) = Z_i f(r)$.
Figure~\ref{fig:gaussian_HJ} shows the results for the apparent radius
distribution, $\Gamma\a(r\a)$.  As one would expect, the effect is to
smear the radius distribution toward smaller values, because some of
the hot Jupiters now appear to be smaller planets.  Less obviously,
the integrated rate of hot Jupiters is also affected.  We calculate
the apparent hot Jupiter rate by integrating the apparent rate density
for $r\a > 8\,R_\oplus$.  If secondaries host any planets at all, the
apparent rate is {\it greater} than the true rate.  For instance, when
hot Jupiters are just as common around secondary stars as single or primary
stars, then $\Lambda_{\rm HJ,a}/\Lambda_{{\rm HJ},0} = 1.23$.
%int_rate_density_vs_radius_model_7_withtext*manyZs.pdf
The apparent rate is higher because a larger number of stars were
searched than are accounted for in the rate calculation.

>>>>

Here, though, we will focus exclusively on problems that arise from
the fact that many stars exist in unresolved binary systems.  We will
also focus on the errors in planet occurrence as a function of radius,
rather than orbital period. This is because when more than one transit
is detected (as is usually required by the surveyors), the orbital
periods can be measured without ambiguity regardless of whether the
host star is single or one member of a binary.  Even with this narrow
focus, there are numerous sources of error.  All three of the
quantities in Equation~\ref{eq:occ_rate_simple} are biased:
\begin{enumerate}
%    
    \item The number of detected planets, $N_{\det}$, is actually the
      number of detected planets that {\it appear} to have size $r$,
      orbiting stars that {\it appear} to have mass $M$.  Whenever the
      planet-hosting star is part of a binary,
%    
    \begin{itemize}
        \item the planet's size could be misclassified because of the
          reduction in the amplitude of the photometric signal;
%        
        \item the host star's properties could be misclassified
          because its light is combined with a second star of a
          different spectral type.
%        
    \end{itemize}
%    
    \item The number of searchable stars, $N_\star$, is biased
%    
    \begin{itemize}
%        
        \item toward lower values, because it does not include all of
          the secondary stars that were inadvertently searched for
          transiting planets;
%        
        \item toward higher values, because some of the stars that
          appeared to be searchable are in fact binaries for which the
          amplitude of the photometric signal would have been reduced
          to an undetectable level.
%        
    \end{itemize}
%    
    \item The transit probability $p_{\rm tra}$ is biased because the
      planet-hosting star could be misclassified.  At fixed orbital
      period, the transit probability is proportional to
      $\rho^{-1/3}$, where $\rho$ is the stellar mean
      density~\citep{Winn2010}.  Therefore, any errors in determining
      the host star's mean density lead to errors in the correction
      for the transit probability.
%    
\end{enumerate}

@@

Since the effects of binarity are most pronounced when the two stellar
components are similar, we begin by considering a universe in which
all stars are identical, with mass $M$, radius $R$, and luminosity
$L$.  Single stars are uniformly distributed in space with a number
density of $n\s$ stars per cubic parsec, and binary systems are
uniformly distributed with number density $n\b$.  In this scenario,
stars are never misclassified because the combined light of a binary
has the same color and spectrum as a single star.  We further assume
that all planets have the same radius, $r$, and occur around single
stars and members of binaries at the same rate, $\Lambda(r)$.

Our naive observers conduct a transit survey. To calculate the
occurrence rate of planets with radius $r$, they count the number of
detections of signals with amplitude $(r/R)^2$.  Then they identify
all the sources that appear to have been searchable for a signal of
amplitude $(r/R)^2$.  Here and throughout the rest of this paper, we
assume that the limiting source of noise is the photon-counting noise
from the source, i.e., $\sigma \propto 1/\sqrt{F}$, where
where $F$ is the total flux of the source.
Thus the observers determine the minimum $F_0$ for which detection would have been
possible, and count the number of sources with $F>F_0$. This will
include all the single stars out to a maximum distance
\begin{equation}
  d_0 = \sqrt{\frac{L}{4\pi F_0}}.
  \label{eq:dmax}
\end{equation}
Since binaries are twice as luminous, the condition $F>F_0$ will
include binaries out to the larger distance of $d_0\sqrt{2}$.
None of the stars in binaries will
appear to have a planet of radius $r$, because of the dilution of
the transit signal.  Thus, the {\it apparent} occurrence rate $\Lambda\a$ of
planets of radius $r$ is
\begin{equation}
    \Lambda\a(r) = 
        \frac{\Lambda(r) \, n\s d_0^3}
        {n\s d_0^3 + n\b (d_0\sqrt{2})^3}.
    \label{eq:rate_at_rp}
\end{equation}
This apparent rate is larger than the true rate by a factor
\begin{equation}
    \frac{\Lambda\a(r)}{\Lambda(r)} = 
        \frac{1}{1 + 2^{3/2}(n\b/n\s)}.
    \label{eq:correction_rp}
\end{equation}

The observers will also detect some transiting planets around stars
with binary companions.  The amplitude of these signals is
\begin{equation}
  \frac{L(r/R)^2}{L + L} = \frac{1}{2}\left(\frac{r}{R}\right)^2 =
  \left(\frac{r/\sqrt{2}}{R}\right)^2,
    \label{eq:delta_obs_general} 
\end{equation}
leading the astronomers to believe they have discovered a population
of planets with radius $r/\sqrt{2}$.  To calculate the corresponding
occurrence rate, they count the stars for which this type of signal
would have been detectable.  The limiting flux for detection in this
case is $4F_0$, because the signal amplitude has been reduced by a
factor of two and the noise level must also be reduced by a factor of
two.  The condition $F>4F_0$ is met for single stars within a distance
$d_0/2$, and binaries within a distance $d_0\sqrt{2}/2$.  Therefore,
the observers will calculate the occurrence rate of this new type of
planet to be
\begin{equation}
    \Lambda\a\left( \frac{r}{\sqrt{2}} \right) = 
        \frac{2\Lambda(r) \, n\b (d_0\sqrt{2}/2)^3}
        {n\s (d_0/2)^3 + n\b (d_0\sqrt{2}/2)^3}
    =
    \frac{2\Lambda(r)\cdot2^{3/2}(n\b/n\s)}{1 + 2^{3/2} (n\b/n\s)}.
    \label{eq:correction_diluted_rp}
\end{equation}
Figure~\ref{fig:model_1_volumes} illustrates the
volumes enclosing the apparently searchable single and binary stars.

@@

From within this sample of apparently searchable sources, the
observers sum the number of detections ${\rm d}N_{\rm det}$ of planets
with an apparent radius between $r\a$ and $r\a + dr\a$.  There will be
three contributions to this sum,
\begin{align}
  {\rm d}N_{{\rm det},0} &= 
      \frac{N_\star(r\a) p_{\rm tra}}{1+\mu}
      \Gamma_0(r\a)\mathrm{d}r\a,
  \label{eq:n0} \\
  {\rm d}N_{{\rm det},1} &= 
      \frac{\mu N_\star(r\a) p_{\rm tra}}{1+\mu} 
      \Gamma_1(\sqrt{2}r\a)\mathrm{d}(\sqrt{2}r\a),
	\label{eq:n1} \\
  {\rm d}N_{{\rm det},2} &=
      \frac{\mu N_\star(r\a) p_{\rm tra}}{1+\mu} 
      \Gamma_2(\sqrt{2}r\a)\mathrm{d}(\sqrt{2}r\a).
	\label{eq:n2}
\end{align}
where here and elsewhere, the subscript 0 refers to single stars, 1
refers to primary stars of binaries, and 2 refers to secondary stars
of binaries. The apparent occurrence rate density is therefore
\begin{equation}
  \Gamma\a(r\a) = \frac{{\rm d}N_{\rm det}}{{\rm d}r\a} \frac{1}{N_\star(r\a) p_{\rm tra}} =
  \frac{\Gamma_0(r\a)}{1+\mu} +
  \frac{\mu\sqrt{2}\,\Gamma_1(\sqrt{2}r\a)}{1+\mu} +
  \frac{\mu\sqrt{2}\,\Gamma_2(\sqrt{2}r\a)}{1+\mu}.
	\label{eq:formula_twin_binary}
\end{equation}
Our earlier results, Equations~\ref{eq:rate_at_rp}
and~\ref{eq:correction_diluted_rp}, are recovered by integrating this formula
over $r$ after inserting
\begin{equation}
  \Gamma_i(r) = \Lambda(r\p)\, \hat{\delta}(r-r\p),~~{\rm for}\ i\in\{0,1,2\}
\end{equation}
where $\hat{\delta}$ is the Dirac delta function, and $r\p$ is the
single planet size considered in Section~\ref{sec:model_1}.

@@

The most obvious aspect of the results is that the occurrence rate of
planets smaller than the breakpoint radius of 2\,$R_\oplus$ is now
substantially overestimated.  If secondary stars have the same planet
population as the single stars, then the occurrence rate density of
small planets is overestimated by $50\%$.  The magnitude of the
systematic error decreases if there are fewer planets around
secondaries; if secondaries host no planets at all, the occurrence
rate is overestimated by only $10\%$.

The effects on larger planets are not as substantial.  To be
quantitative we consider the occurrence rate of giant planets with
$r>8\,R_\oplus$, by integrating the occurrence rate density displayed
in Figure~\ref{fig:occ_rate_model_3_log}:
\begin{equation}
    \Lambda_{\rm giant,a} =
      \int_{8\,R_\oplus}^{\infty} \Gamma\a(r\a)\,{\rm d}r\a,~~{\rm
      and}~~
    \Lambda_{\rm giant,0} =
      \int_{8\,R_\oplus}^{\infty} \Gamma_0(r)\,{\rm d}r.
\end{equation}
In this case the occurrence rate density is underestimated.  The
difference between the apparent and true rates is largest when the
secondary stars do not host any planets, and has a magnitude
$\Lambda_{{\rm giant},0}/\Lambda_{\rm giant,a} = 1.13$.
%occ_rate_vs_radius_model_3_withtext_Zsub2_0.00_rpu_22.5.pdf
If instead the secondary stars have half as many giant planets as single
stars, then the factor is reduced from 1.13 to 1.06.

@@

The occurrence rate calculations that led to this discovery did not
carefully account for unresolved binaries.  Intuitively we expect
unresolved binaries to have a blurring effect, filling in any gaps in
the radius distribution and making them appear less empty than in
reality.  To investigate the quantitative effects we make identical
assumptions as in Section~\ref{sec:model_3}, except that the radius
distribution is assumed to have a complete absence of planets with
sizes between 1.5 and 2\,$R_\oplus$:
\begin{align}
    f(r)
    &\propto
    \left.
    \begin{cases}
        r^\gamma & \text{for } r\geq 2\,R_\oplus, \\
        0 & \text{for } 1.5\,R_\oplus < r < 2\,R_\oplus, \\
        {\rm constant} & \text{for } r\leq 1.5\,R_\oplus.
    \end{cases}
    \right.
    \label{eq:model4_radius_distribution}
\end{align}
Figure~\ref{fig:model_4} shows the resulting true and apparent occurrence
rate densities.
For the case $Z_2/Z_0=1$, unresolved binaries reduces the contrast
of the gap by nearly a factor of two, while also producing
spurious features for apparent sizes beneath 1.5\,$R_\oplus$.
Thus the gap identified by \citet{fulton_california-_2017} may be
even more devoid of planets than it appears.

@@

The planet population may include some special members with a distinct
radius distribution.  For example in the recent study
by~\citet{petigura_CKS_2017}, hot Jupiters appear as an island in
period-radius space, rather than as a component of a power-law
distribution extending to smaller planets and longer periods.
For this reason we test the effects of unresolved binaries
on a Gaussian radius distribution,
\begin{equation}
    f(r) \propto \exp \left[ -\frac{(r-\bar{r})^2}{2\sigma_r^2} \right],
\end{equation}
with $\bar{r} = 14\,R_\oplus$ and $\sigma_r = 2\,R_\oplus$.  As before,
we allow for different normalizations of the planet populations around
single, primary, and secondary stars: $\Gamma_i(r) = Z_i f(r)$.
Figure~\ref{fig:gaussian_HJ} shows the results for the apparent radius
distribution, $\Gamma\a(r\a)$.  As one would expect, the effect is to
smear the radius distribution toward smaller values, because some of
the hot Jupiters now appear to be smaller planets.  Less obviously,
the integrated rate of hot Jupiters is also affected.  We calculate
the apparent hot Jupiter rate $\Lambda_{\rm HJ,a}$ by integrating the
apparent rate density for $r\a > 8\,R_\oplus$.  If secondaries host
any planets at all, the apparent rate is {\it greater} than the true
rate.  For instance, when hot Jupiters are just as common around
secondary stars as single or primary stars, then $\Lambda_{\rm
HJ,a}/\Lambda_{{\rm HJ},0} = 1.23$.
%int_rate_density_vs_radius_model_7_withtext*manyZs.pdf
The apparent rate is higher because a larger number of stars were
searched than are accounted for in the rate calculation.
